%--------------------------------------------------------------------------
\chapter{Conclusion} \label{ch:conclusion}
%--------------------------------------------------------------------------


This dissertation investigated the acoustics of voice quality in Santiago Laxopa Zapotec and its interactions with tone. There were four main questions that this dissertation addressed: (i) does the recently proposed residual H1* acoustic measure more effective captures the phonation contrasts between Santiago Laxopa Zapotec's four phonations than the traditional H1*$-$H2* acoustic measure; (ii) how is the acoustic landscape of voice quality in Santiago Laxopa Zapotec structured; (iii) which acoustic measures most effectively capture and classify the voice quality contrasts; (iv) and how do those acoustic measures help explain Santiago Laxopa Zapotec's laryngeal complexity? 

%------------------------------------------
\section{Summary of findings}\label{sec:summary_of_findings}
%-------------------------------------------

This dissertation showed in Chapter~\ref{ch:residual_h1}, that \posscitet{chaiH1H2AcousticMeasure2022} residual H1* is a more effective acoustic measure for capturing the phonation contrasts in Santiago Laxopa Zapotec than the traditional H1*$-$H2* measure. Furthermore, this measure was also shown to play an important role in defining the phonation's acoustic landscape in Santiago Laxopa Zapotec, as demonstrated in Chapter~\ref{ch:acousticlandscape}. Additionally, Chapter~\ref{ch:revealing_trees} showed that residual H1* also was one of the important acoustic measures for classifying the phonation contrasts in Santiago Laxopa Zapotec according to a Random Forest model. These results contribute to validating \posscitet{chaiH1H2AcousticMeasure2022} residual H1* measure as a reliable acoustic measure for phonation contrasts as does recent work by \citet{chaiPhoneticsGlottalizedPhonations2023} and \citet{garellekMarginsPhonologyPhonetics2025}. 

Chapter~\ref{ch:acousticlandscape} demonstrated that SLZ's phonation largely occupies a three-dimensional acoustic landscape defined by spectral slope and harmonic structure/noise. These findings are consistent with the findings from \citet{keatingCrosslanguageAcousticSpace2023}. In both studies, these first two dimensions are correlated with spectral slope and harmonic structure. The study presented in Chapter~\ref{ch:acousticlandscape} showed that SLZ additionally needed a third dimension, also correlated with spectral slope, to fully capture phonation's dimensionality. The results of these studies suggest that phonation is primarily characterized by just spectral slope and harmonic structure. Furthermore, the results of this study suggest that when higher dimensionality is needed, spectral slope or harmonic structure/noise are also correlated with these dimensions.

Chapter~\ref{ch:revealing_trees} demonstrated that SLZ's phonation relies on a small set of acoustic measures to effectively classify the phonation contrasts. The Random Forest model presented in this chapter showed that many of the same measures that were correlated with the dimensions of the acoustic landscape were also important for classification. The results of the Random Forest analysis showed that the most important acoustic measures for classification were: (i) duration, (ii) A1*, (iii) H1*$-$A1*, (iv) residual H1*, (v) HNR $<$ 1500 Hz, and (vi) Strength of Excitation.These findings suggest that the phonation contrasts in SLZ are not only characterized by spectral slope and harmonic structure/noise, but also by the duration of the vowel and the strength of excitation. These results contribute to our understanding of how phonation contrasts can be effectively classified using a small set of acoustic measures.

The previous chapters were important in establishing which acoustic measures needed to be consulted for Chapter~\ref{ch:testing_lc}'s analysis of how phonation interacts with tone in SLZ. This chapter investigated these interaction through generalized additive mixed model analysis of \textit{f}0, HNR $<$ 1500 Hz, and Strength of Excitation; revealing two facts about how laryngeal complexity manifests itself in SLZ. First, the results of this chapter showed that nonmodal phonation is not strongly articulated in SLZ. This was demonstrated by the very small, though statistically significant, differences in Strength of Excitation between the modal and nonmodal phonations. If nonmodal phonation were strongly articulated in SLZ, we would expect to see larger differences in Strength of Excitation between the modal and nonmodal phonations. This first fact confirms \posscitet{herrerazendejasAmuzgoZapotecTwo2000} claim that Zapotec languages do not articulate their nonmodal phonation strongly. Second, the results of this chapter also showed that contrary to \posscitet{herrerazendejasAmuzgoZapotecTwo2000} claims that Zapotec languages lack phasing of modal and nonmodal phonation, SLZ does exhibit phasing of modal and nonmodal phonation. These results demonstrate that \citet{silvermanLaryngealComplexityOtomanguean1997,silvermanPhasingRecoverability1997} was correct in their assertion that laryngeal complexity primarily exhibits itself through phasing of modal and nonmodal phonation. The results additionally showed that despite lacking strong articulation of nonmodal phonation, SLZ does exhibit phasing of modal and nonmodal phonation which runs counter to \possciteauthor{silvermanLaryngealComplexityOtomanguean1997} claims that either you exhibit phasing with strong articulation or you lack phasing and weakly articulate nonmodal phonation.

These finding contribute to our understanding of phonation and laryngeal complexity; however, these findings also raise several questions important questions that require further investigation. These questions will be discussed in the following sections. 

%-------------------------------------------
\section{Future directions}\label{sec:future_directions}
%-------------------------------------------

%-------------------------------------------
% \subsection{Perception of phonation and tone}\label{sec:perception}
%-------------------------------------------

Despite the growing number of perceptual studies on phonation or tone, perception research into laryngeal complexity is lacking. This dissertation has shown shown that phonation is primarily analyzed through spectral slope and the harmonic structure/noise of the signal, which is correlated to the open quotient of the glottis and the periodicity of the signal. However, it is still unclear how these tones and phonation contrasts are perceived by speakers of SLZ. 

Research has shown that even though these acoustic measures are effective for classifying phonation contrasts, they do not always correlate with how speakers perceive these contrasts. For example, \citet{hillenbrandRoleF0Amplitude1996} and \citet{gerfenProductionPerceptionLaryngealized2005} found that speakers of American English and Mixtec, respectively, primarily relied on changes in \textit{f}0 and amplitude to perceive intervocalic laryngealization. Furthermore, checked vowels have as their percepts a short duration and falling \textit{f}0 contour are more important cues than late-phased glottalization \citep{brunelleTonePerceptionSgaw2011,chaiPhoneticsPhonologyChecked2022,chaiPerceptionCheckedRearticulated2025,garellekVoiceQualityTone2013}. \citet{klattAnalysisSynthesisPerception1990} found that aspiration noise was the important cue for distinguishing breathy and modal phonation apart in English; while \citet{hillenbrandAcousticCorrelatesBreathy1996} found that listeners primarily relied on the cepstral peak prominence to distinguish breathy and modal phonation. Because of phonation's multidimensionality, it is likely that listeners may rely on different acoustic cues to perceive phonation contrasts \citep{espositoEffectsLinguisticExperience2010,kreimanUnifiedTheoryVoice2014,kreimanValidatingPsychoacousticModel2021}. Besides \citet{gerfenProductionPerceptionLaryngealized2005}, these studies have primarily investigated languages that lack laryngeal complexity. Therefore, it is unclear how speakers of SLZ perceive the phonation contrasts in their language. This is an important question to address because it will help us understand how phonation contrasts are perceived in languages with laryngeal complexity.

%-------------------------------------------
% \subsection{Typological implications}\label{sec:typological_implications}
%-------------------------------------------

%-------------------------------------------
% \subsection{The phonology of laryngeal complexity}\label{sec:phonology_lc}
%-------------------------------------------



%--------------------------------------------------------------------------
% \section{The Laryngeal Articulator Model}\label{sec:lam}
% %--------------------------------------------------------------------------

% \citet{eslingThereAreNo2005,eslingVoiceQualityLaryngeal2019,moisikPhonologicalPotentialsLower2021,moisikMultimodalImagingGlottal2015,moisikModelingBiomechanicalInfluence2014}

% \begin{figure}
%     \centering
%     \begin{tikzpicture}
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (tra) at (0,0) {tra};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (tfr) at (-2,-1.5) {tfr};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (tre) at (2,-1.5) {tre};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (tdb) at (0,-3) {tdb};

%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (epc) at (0,-6) {epc};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (vfo) at (-2,-7.5) {vfo};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (vfc) at (2,-7.5) {vfc};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (epv) at (0,-9) {epv};

%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (lower_lx) at (-3.5,-6) {↓lx};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (raised_lx) at (3.5,-6) {↑lx};

%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (Lf0) at (-2,-10.5) {Lf0};
%         \node[draw,circle,minimum size=1cm,inner sep=0pt] (Hf0) at (2,-10.5) {Hf0};
        
%         % Anti-syngeristic
%         \draw[<->, dotted, line width=.5mm] (tra.west) to[out=180,in=90] (tfr.north);
%         \draw[<->,dotted, line width=.5mm] (tra.east) to[out=0,in=90] (tre.north);
%         \draw[<->,dotted, line width=.5mm] (tfr.east) to (tre.west);
%         \draw[<->,dotted, line width=.5mm] (tra.south) to (tdb.north);
%         \draw[<->,dotted, line width=.5mm] (tfr.south) to[in=180, out=-90] (epc.west);
%         \draw[<->,dotted, line width=.5mm] (lower_lx.north) to[in=140,out=40] (raised_lx.north);
%         \draw[<->,dotted, line width=.5mm] (lower_lx.east) to (epc.west);
%         \draw[<->,dotted, line width=.5mm] (epc.west) to[in=90,out=180] (vfo.north);
%         \draw[<->,dotted, line width=.5mm] (vfo.east) to[in=180,out=0] (vfc.west);
%         \draw[<->,dotted, line width=.5mm] (vfc.south) to[in=0,out=-90] (epv.east);
%         \draw[<->,dotted, line width=.5mm] (epc.south) to[out=-90, in=90] (Hf0.north);
%         \draw[<->,dotted, line width=.5mm] (Hf0.south) to[out=-140, in=-40] (Lf0.south);
%         \draw[<->,dotted, line width=.5mm] (epv.south) to[out=-90, in=180] (Hf0.west);

%         % Syngeristic
%         \draw[<->, line width=.5mm] (tdb.west) to[out=180, in=-90] (tfr.south);
%         \draw[<->, line width=.5mm] (tdb.east) to[out=0, in=-90] (tre.south);
%         \draw[<->, line width=.5mm] (epc.north) to (tdb.south);
%         \draw[<->, line width=.5mm] (epc.south) to (epv.north);
%         \draw[<->, line width=.5mm] (epc.east) to[in=-90, out=0] (tre.south);
%         \draw[<->, line width=.5mm] (lower_lx.north) to[in=-135, out=90] (tfr.west);
%         \draw[<->, line width=.5mm] (raised_lx.north) to[in=-45, out=90] (tre.east);
%         \draw[<->, line width=.5mm] (Lf0.west) to[in=-90, out =135] (lower_lx.south);
%         \draw[<->, line width=.5mm] (raised_lx.west) to (epc.east);
%         \draw[<->, line width=.5mm] (Hf0.east) to[in=-90, out=45] (raised_lx.south);
%         \draw[<->, line width=.5mm] (epv.south) to[out=-90, in=0] (Lf0.east);
%         \draw[<->, line width=.5mm] (epc.east) to[out=0, in=90] (vfc.north);
%         \draw[<->, line width=.5mm] (vfo.south) to[out=-90, in=180] (epv.west);
%         \draw[<->, line width=.5mm] (vfc.south) to (Hf0.north);
%         \draw[<->, line width=.5mm] (vfo.south) to (Lf0.north);
%         \draw[<->, line width=.5mm] (epc.south) to[out=-90,in=90] (Lf0.north);
%         \draw[<->, line width=.5mm] (lower_lx.south) to[out=-90,in=180] (vfo.west);
%         \draw[<->, line width=.5mm] (raised_lx.south) to[out=-90,in=0] (vfc.east);

%         % Curly bracket with text
%         \draw[decorate,decoration={brace,amplitude=10pt,raise=5pt},thick] (4,0) -- (4,-4) node[midway,xshift=15pt,right=5pt] {\parbox{2cm}{vowel\\quality}};
%         \draw[decorate,decoration={brace,amplitude=10pt,raise=5pt},thick] (4,-4) -- (4,-9.5) node[midway,xshift=15pt,right=5pt] {\parbox{2cm}{phonatory\\quality}};
%         \draw[decorate,decoration={brace,amplitude=10pt,raise=5pt},thick] (4,-9.5) -- (4,-11) node[midway,xshift=15pt,right=5pt] {\parbox{2cm}{tonal\\quality}};
%         \draw[decorate,decoration={brace,amplitude=10pt,raise=5pt},thick] (7,-2) -- (7,-11) node[midway,xshift=15pt,right=5pt] {\parbox{2cm}{voice\\quality}};

        
%     \end{tikzpicture}
%     \caption{The Laryngeal Articulator Model from \citet{eslingVoiceQualityLaryngeal2019}. This model shows the interactions between the laryngeal articulators (labeled circles). Syngeristic interactions are shown with solid lines, while anti-syngeristic interactions are shown with dotted lines.}
%     \label{fig:tra}
% \end{figure}

% %--------------------------------------------------------------------------
% \section{Modeling laryngeal complexity}\label{sec:modeling_lc}
% %--------------------------------------------------------------------------

% %--------------------------------------------------------------------------
% \section{Alternative accounts}\label{sec:alternative_accounts}
% %--------------------------------------------------------------------------

% %--------------------------------------------------------------------------
% \subsection{Articulatory Phonology account}\label{sec:ap_account}
% %--------------------------------------------------------------------------

% %--------------------------------------------------------------------------
% \subsection{Q-theory account}\label{sec:q_theory_account}
% %--------------------------------------------------------------------------

% %--------------------------------------------------------------------------
% \subsection{Radical CV Phonology account}\label{sec:rcv_account}
% %--------------------------------------------------------------------------